---
title: "Notes"
author: "Brandon Blanchette"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Notes for GAMs course

# Day 1

Exercises : 

Start will belgium.rmd then do heart-attack.rmd



# Day 2 

Exercises :

Start with co2_south-pole-example then larks then linnets

# Day 3

Slides : 

## 8
Test to figure out if a non linear model is suitable for the data but in Gavin's opinion we should not have to do this because the penalty will not allow over wiggliness.

```{r}
library(tidyverse)
library(mgcv)
n <- 100
set.seed(2)
df <- tibble(x = runif(n),
             y = x + x^2 * 0.2 + rnorm(n) * 0.1)
model <- gam(y ~ x + s(x, m = c(2, 0)),
             data = df, method = "REML")
model2 <- gam(y ~ s(x),
             data = df, method = "REML")

summary(model)
summary(model2)
```


## 11
not all of the smooths can be used to estimate the second derivative as the penalty makes the second derivative as small as possible



## 21

Why the thin plate spline doesn't work when the scale on x and z is different but the tensor product is much better for this case


## 25 
Last line : make sure that the data in place of "f" is a factor if not you will get a cryptic error message


## 37

we should look to see if there's a difference between k' and edf and if k-index is low (unmodelled wigliness) 

if the k index is too low it means that the values near each other aren't similar we want the k-index to be greater than one or at least close (if its close check the p-value if not don't worry about the p-value too much)

## 38
s(x2) doesnt have enough degrees of freedom now that s(x1) is figured out and has enough degrees of freedom

s(x2) edf and k' are good (there is enough head room in terms of k for the)

## 40 

we need more degrees of freedom to force the curve to go flat (slope is changing faster more often so we need more degrees of freedom)

## 41

there a 2 scenarios where this test will fail

## 42 
k-index is not the same here as we saw before

## 43

fitting gam model to residuals 

if we had the basis size correct the edfs should be 1 here they are not

## 45

model diagnostics we need to do if we are to validate any model

## 47

3. can help us diagnose non constant variance

##48

here is what happens when we have the correct model

rep = 500 : simulates 500 new data sets from the model using the fitted values

if your data looks like a typical example from the model then your model is good

you want all your points to be on line line in the first graph you can expect some variation though

second graph : if there is an increase in spread if we increase the mean of the residuals then we have the wrong mean residuals relationship **It should be constant**

## 49

example of bad model fit overdispersion 

plot 3 is too fat in the tails and we can see the same thing in the first plot 

## 51

do the same thing as before but using ggplot

## 55

here we have the wrong family so we have overdispersion and we are testing for that 

our dispersion should be 1 so its very overdispersed p-value is significant so does not support the null hypothesis that it is correctly dispersed

## 56
no sign that the data and model are different in terms of their dispersions

## 57

takes our fitted model and simulates residuals and we can do other tests

## 58 

KS test tells us if we have the correct distribution significant means we dont have the correct family

## 59 

this is what good model diagnostics looks like

## 61

we should be really careful about trying to get something really parsimonious

##  67

3rd arrow adding several variables that have no effect 

very noisy situation 

## 68 

edf for all the terms that have no effect have all been shrunk to 0 or almost 0 so are basically all out of the model 

## 69

scales = fixed gives us all the same scales for all the plots

## 74
expected value of temperature : E

F(TOD) = water temperature will vary depending on time of day it was sampled

f(DOY) = water temperature will vary throughout the year

f(Year) = long term water temperature variation 


## 75
f(x,y,TOD) effect of time of day might be different depending on location

##78 
we use ti here so it doesnt include the main effects multiple times in the model ti() does not include the main effect of a variable so we can use it s() to add the main effects and ti() to add the interactions between the variables

knots upper limit set to 366.5 to get predictions for years with leap years 

we should make sure we're using the same basis functions for the main effects and the interactions for the same variable **for example bs = "ds" for LONGITUDE LATITUDE**

## 79
reasonable null model that just has the effect of climate change

## 81 

clearly the spatial model is much better than the null model

## 83

probably a problem with the data in the figure in the top right the temperature would not increase that fast and settle back down that fast

## 84

with the development version of gratia you can make theses plots 

DOY seems to be the variable with the most effect here

time of day seems to have the smallest effect since it has the smallest range

## 85 

data slice TOD = 12 and DOY 180

so we're fixing the time of day and the day of year so that we can predict the spatial effects

## 89 

lets pick a point of the bay and we want to see the trends at that point at different days of the year

## 90

seems to be a trend in the fall where the temperature is increasing 


# Day 4

Slides : 

## 11 

you should probably use this if it is available to use (REML AND ML)

## 19 
we should not be relying on p-values to make decisions according to statisticians we're over relying on them

if the effect is significant and the variable is biologically important then you should keep it regardless of p-value 

## 


