---
title: "Notes"
author: "Brandon Blanchette"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Notes for GAMs course

# Day 1

Exercises : 

Start will belgium.rmd then do heart-attack.rmd



# Day 2 

Exercises :

Start with co2_south-pole-example then larks then linnets

# Day 3

Slides : 

## 8
Test to figure out if a non linear model is suitable for the data but in Gavin's opinion we should not have to do this because the penalty will not allow over wiggliness.

```{r}
library(tidyverse)
library(mgcv)
n <- 100
set.seed(2)
df <- tibble(x = runif(n),
             y = x + x^2 * 0.2 + rnorm(n) * 0.1)
model <- gam(y ~ x + s(x, m = c(2, 0)),
             data = df, method = "REML")
model2 <- gam(y ~ s(x),
             data = df, method = "REML")

summary(model)
summary(model2)
```


## 11
not all of the smooths can be used to estimate the second derivative as the penalty makes the second derivative as small as possible



## 21

Why the thin plate spline doesn't work when the scale on x and z is different but the tensor product is much better for this case


## 25 
Last line : make sure that the data in place of "f" is a factor if not you will get a cryptic error message
