---
title: "Belgium AIDS epidemic"
author: "Gavin Simpson"
date: "10/10/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Description

The data set `belgium-aids.csv` contains the number of cases (`cases`) of AIDS recorded between 1981 and 1993 (`year`) in Belgium

## Packages

Load the following packages
```{r}
rm(list=ls())
```


```{r load-packages}
pkgs <- c("readr", "dplyr", "ggplot2", "here")
vapply(pkgs, library, logical(1L), character.only = TRUE, logical.return = TRUE)
```

## Data

The data are contained in `data/belgium-aids.csv`, and consist of three variables

1.  `year`, the year of each observation
2.  `cases`, the number of AIDS cases per year

Load the data into R

```{r load-data}
belgium <- read_csv(here("data", "belgium-aids.csv"), col_types = "dd")
belgium
```

## Data wrangling

Create a variables `t` (for time) in the `belgium` data frame, which contains the time point of observation by subtracting `1980` from each `year`:

```{r wrangling}
belgium <- belgium %>%
  mutate(t = year - 1980)
```

## Task

1.  Produce a plot of the number of AIDS cases over time

```{r somePlot, echo=FALSE}
ggplot(belgium, aes(x = year, y = cases))+
	geom_point(color = "blue")+
	labs(x = "Time (years)", y = "AIDS cases")
```


2.  Fit a poisson GLM (log link) using the `glm()` to model the number of AIDS cases. Use `t` as the predictor variable

```{r}
m1 <- glm(cases ~ t, data = belgium, family = poisson(link = "log"))
m1
```


3.  Produce the model summary using `summary()`

```{r}
summary(m1)
```


4.  Calculate a quick goodness of fit test for the model using the information provided on the residual deviance

```{r}
pchisq(deviance(m1), df.residual(m1), lower.tail = FALSE)
```


5.  Plot the model diagnostics

```{r}
layout(matrix(1:4, ncol = 2, byrow = TRUE))
plot(m1)
layout(1)
```


6.  Add fitted values of the model as a line on to the plot you created earlier. Add a confidence interval to the fitted value line.

```{r somePlot, echo=FALSE}
pdat <- with(belgium, 
						 tibble(t = seq(min(t),
						 							 max(t),
						 							 length(100))))
pdat <- pdat %>%
  mutate(year = 1980 + t)
pdat

pred <- predict(m1, pdat, type = "link", se.fit = TRUE)

str(pred)

p_link <- predict(m1, pdat, type = "response")
head(p_link)

#inverse link
ilink <- family(m1)$linkinv

#add prediction data and prediction values 
pdat <- pdat %>% 
	bind_cols(data.frame(pred)) %>% 
	mutate(fitted = ilink(fit), 
				 upper = ilink(fit + (1.96 * se.fit)),
				 lower = ilink(fit - (1.96 * se.fit)))

belgium %>%
ggplot(aes(x = year, y = cases)) +
  geom_ribbon(data = pdat,
              aes(ymin = lower, ymax = upper, x = year),
              alpha = 0.2, inherit.aes = FALSE) +
  geom_point() +
  geom_line(data = pdat, aes(y = fitted, x = year)) +
  labs(x = "Creatinine kinase level", y = "Proportion Heart Attack")


```


7.  Looking at the model diagnostics and the plot of the fitted values versus the data, identify a problem with this model. Describe the problem in a few words.

The model does not seem to fit very well with the data at the end.

Residuals vs fitted graph shows a clear problem with the model the points should be spaced evenly. We have the wrong functional form for our model. It looks like we need a ^2 to fix it according to the residuals vs fitted graph

8.  Modify your model to account for the identified problem

    Hint: use polynomials of `t`
    
```{r}
m2 <- glm(cases ~ poly(t, 2), data = belgium, family = poisson(link = "log"))

##  or 

m2b <- update(m1, . ~ . + I(t^2)) #this is less stable numerically so the first option is a bit better

 #this is less stable numerically
```
    

9.  Look at the model summary, model diagnostic plots, and perform the goodness of fit test on the residual deviance. Are there any problems with this model?

```{r}
# model summary
summary(m2)
# quick GOF test
pchisq(deviance(m2), df.residual(m2), lower.tail = FALSE)
# model diagnostics
layout(matrix(1:4, ncol = 2, byrow = TRUE))
plot(m2)
layout(1)

# plots
pred <- predict(m2, newdata = pdat, type = "link", se.fit = TRUE)
ilink <- family(m2)$linkinv
pdat2 <- pdat %>%
  select(-c(fit, se.fit, residual.scale)) %>% # remove m1 link scale fit & std err
  bind_cols(data.frame(pred)) %>%
  mutate(fitted = ilink(fit),
         upper = ilink(fit + (1.96 * se.fit)),
         lower = ilink(fit - (1.96 * se.fit))) %>%
  bind_rows(pdat) %>%
  mutate(model = rep(c("t^2","t"), each = 100))

belgium %>%
ggplot(aes(x = year, y = cases)) +
  geom_ribbon(data = pdat2,
              aes(ymin = lower, ymax = upper, x = year, fill = model),
              alpha = 0.2, inherit.aes = FALSE) +
  geom_point() +
  geom_line(data = pdat2, aes(y = fitted, x = year, colour = model)) +
  labs(x = "Creatinine kinase level", y = "Proportion Heart Attack") +
  scale_fill_discrete(labels = scales::parse_format()) +
  scale_colour_discrete(labels = scales::parse_format())
```


10. Compare the two models using AIC and a generalized likelihood ratio test (GLRT). Which model is to be preferred?

    Hint: the `anova()` function is used to do the GLRT
    
    
    anova(m1, m2, test = "Chisq")

```{r}
anova(m1, m2, test = "Chisq")

anova(m1, m2, test = "LRT")

pchisq(71.446, df = 1, lower.tail = FALSE)

deviance(m1)

deviance(m2)

deviance(m1) - deviance(m2)
```

