---
title: "Belgium AIDS epidemic"
author: "Gavin Simpson"
date: "10/10/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Description

The data set `belgium-aids.csv` contains the number of cases (`cases`) of AIDS recorded between 1981 and 1993 (`year`) in Belgium

## Packages

Load the following packages

```{r load-packages}
pkgs <- c("readr", "dplyr", "ggplot2", "here")
vapply(pkgs, library, logical(1L), character.only = TRUE, logical.return = TRUE)
```

## Data

The data are contained in `data/belgium-aids.csv`, and consist of three variables

1.  `year`, the year of each observation
2.  `cases`, the number of AIDS cases per year

Load the data into R

```{r load-data}
belgium <- read_csv(here("data", "belgium-aids.csv"), col_types = "dd")
belgium
```

## Data wrangling

Create a variables `t` (for time) in the `belgium` data frame, which contains the time point of observation by subtracting `1980` from each `year`:

```{r wrangling}
belgium <- belgium %>%
  mutate(t = year - 1980)
```

## Task

1.  Produce a plot of the number of AIDS cases over time

```{r somePlot, echo=FALSE}
ggplot(belgium, aes(x = year, y = cases))+
	geom_point(color = "blue")+
	labs(x = "Time (years)", y = "AIDS cases")
```


2.  Fit a poisson GLM (log link) using the `glm()` to model the number of AIDS cases. Use `t` as the predictor variable

```{r}
m1 <- glm(cases ~ t, data = belgium, family = poisson(link = "log"))
m1
```


3.  Produce the model summary using `summary()`

```{r}
summary(m1)
```


4.  Calculate a quick goodness of fit test for the model using the information provided on the residual deviance

```{r}
pchisq(deviance(m1), df.residual(m1), lower.tail = FALSE)
```


5.  Plot the model diagnostics

```{r}
layout(matrix(1:4, ncol = 2, byrow = TRUE))
plot(m1)
layout(1)
```


6.  Add fitted values of the model as a line on to the plot you created earlier. Add a confidence interval to the fitted value line.

```{r}
pdat <- with(belgium, 
						 tibble(t = seq(min(t),
						 							 max(t),
						 							 length(100))))
pdat <- pdat %>%
  mutate(year = 1980 + t)
pdat

pred <- predict(m1, pdat, type = "link", se.fit = TRUE)

str(pred)

p_link <- predict(m1, pdat, type = "response")
head(p_link)

#inverse link
ilink <- family(m1)$linkinv

#add prediction data and prediction values 
pdat <- pdat %>% 
	bind_cols(data.frame(pred)) %>% 
	mutate(fitted = ilink(fit), 
				 upper = ilink(fit + (1.96 * se.fit)),
				 lower = ilink(fit - (1.96 * se.fit)))

belgium %>%
ggplot(aes(x = year, y = cases)) +
  geom_ribbon(data = pdat,
              aes(ymin = lower, ymax = upper, x = year),
              alpha = 0.2, inherit.aes = FALSE) +
  geom_point() +
  geom_line(data = pdat, aes(y = fitted, x = year)) +
  labs(x = "Creatinine kinase level", y = "Proportion Heart Attack")


```


7.  Looking at the model diagnostics and the plot of the fitted values versus the data, identify a problem with this model. Describe the problem in a few words.

The model does not seem to fit very well with the data at the end

8.  Modify your model to account for the identified problem

    Hint: use polynomials of `t`
    
```{r}
m2 <- glm(cases ~ poly(t, 2), data = belgium, family = poisson(link = "log"))

#  or 

m2a <- update(m1, . ~ . - t + poly(t, 2))
```
    

9.  Look at the model summary, model diagnostic plots, and perform the goodness of fit test on the residual deviance. Are there any problems with this model?

10. Compare the two models using AIC and a generalized likelihood ratio test (GLRT). Which model is to be preferred?

    Hint: the `anova()` function is used to do the GLRT
