---
title: "Heart attacks & creatinine kinase"
author: "Gavin Simpson"
date: "10/10/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Description

Patients who experience a heart attack tend to do better if the condition is diagnosed early. Smith (1967) describes a study that looked into using the levels of the enzyme *creatinine kinase* (`ck`) in patients' blood. The creatinine kinase level was measured in the blood of 360 patients suspected of suffering a heart attack. Whether each patient had suffered a heart attack was determined later with traditional medical care and tests.

## Packages

Load the following packages

```{r load-packages}
pkgs <- c("readr", "dplyr", "ggplot2", "here")
vapply(pkgs, library, logical(1L), character.only = TRUE, logical.return = TRUE)
```

## Data

The data are contained in `data/heart-attack.csv`, and consist of three variables

1. `ck`, the creatinine kinase level for each patient. The original paper binned patients into groups of CK levels. Here we use the mid points of the bins,
2. `attack`, the number of patients who experienced a heart attack
3. `ok`, the number of patients who did not experience a heart attack

Load the data into R

```{r load-data}
ck <- read_csv(here("data", "heart-attack.csv"), col_types = "ddd")
ck
```

## Data wrangling

Create a variables in `ck` for:

1. `total` --- the total number of patients in each `ck` bin, and
2. `proportion` --- the proportion of the total number of patients that suffered a heart attack

```{r wrangling}
ck <- ck %>%
  mutate(total = attack + ok,
         proportion = attack / total)
```

## Task

1. Produce a plot of the proportion of patients who suffered a heart attack vs the creatinine kinase level in the patient's blood

```{r}
ck %>% 
	ggplot(aes( x = ck, y = proportion))+
	geom_point()+
	geom_smooth(method = "loess") +
  labs(x = "Creatinine kinase level", y = "Proportion Heart Attack")
```


2. Fit a binomial GLM (logit link) using the `glm()` to model the proportion of patients who suffered a heart attack as a function of the creatinine kinase level in the patient's blood

    Hint: If you're not sure how to specify the response for the GLM, read the *Details* section of `?glm`. You can use either the

        * two-column matrix version or,
        * the proportion as response but specify total counts via the `weights` argument
        
```{r}
m_ck <-  glm(proportion ~ ck, data = ck, family = binomial("logit"),
						 weights = total)
```
        

3. Produce the model summary using `summary()`

```{r}
summary(m_ck)
```

    Is the effect of `ck` on the proportion of patients with heart attacks statistically significant?
    
    **yes**
    
    What is the proportion of deviance explained?
    

```{r deviance explained}
summ <- summary(m_ck)
null_dev <- summ$null.deviance
(null_dev - deviance(m_ck)) / null_dev
```

    

4. Calculate a quick goodness of fit test for the model using the information provided on the residual deviance

```{r gof-test}
pchisq(deviance(m_ck), df.residual(m_ck), lower.tail = FALSE)


```
very small probability that we would have a deviance this large and thus the model is poorly fitted for the data


5. Plot the model diagnostics

```{r model diagnostics} 

#alternative to layout code
op <- par(mfrow = c(2,2))
plot(m_ck)
par(op)
```


6. Add fitted values of the model as a line on to the plot you created earlier

```{r}
ck <- ck %>%
  mutate(fitted_1 = fitted(m_ck))
ggplot(ck, aes(x = ck, y = proportion)) +
  geom_point() +
  geom_line(aes(y = fitted_1)) +
  labs(x = "Creatinine kinase level", y = "Proportion Heart Attack")
```
```{r}
pdat <- with(ck, tibble(ck = seq(min(ck), max(ck), length = 100)))
pred <- predict(m_ck, newdata = pdat, type = "link", se.fit = TRUE)
ilink <- family(m_ck)$linkinv
pdat <- pdat %>%
  bind_cols(data.frame(pred)) %>%
  mutate(fitted = ilink(fit),
         upper = ilink(fit + (1.96 * se.fit)),
         lower = ilink(fit - (1.96 * se.fit)))

ggplot(ck, aes(x = ck, y = proportion)) +
  geom_ribbon(data = pdat,
              aes(ymin = lower, ymax = upper, x = ck),
              alpha = 0.2, inherit.aes = FALSE) +
  geom_point() +
  geom_line(data = pdat, aes(y = fitted, x = ck)) +
  labs(x = "Creatinine kinase level", y = "Proportion Heart Attack")
```


7. Looking at the model diagnostics and the plot of the fitted values versus the data, identify a problem with this model. Describe the problem in a few words.

It looks as though the polynomial of the model is incorrect as the data on the residuals vs fitted plot have a high deviance for points 7 8 and 9 

8. Modify your model to account for the identified problem


    Hint: use ploynomials of `ck`
    
```{r}
m2_ck <- glm(cbind(attack, ok) ~ poly(ck, 3), data = ck, family = binomial(link = "logit"))
```
  

9. Look at the model summary, model diagnostic plots, and perform the goodness of fit test on the residual deviance. Are there any problems with this model?

```{r}
summary(m2_ck)

op2 <- par(mfrow = c(2,2))
plot(m2_ck)
par(op2)

pchisq(deviance(m2_ck), df.residual(m2_ck), lower.tail = FALSE)
```
good level of fit here could be slightly overfitted here


10. Compare the two models using AIC and a generalized likelihood ratio test (GLRT). Which model is to be preferred?

    Hint: the `anova()` function is used to do the GLRT
    
```{r}
AIC(m_ck, m2_ck)

anova(m_ck, m2_ck, test = "LRT")
```
```{r}
pdat <- with(ck, tibble(ck = seq(min(ck), max(ck), length = 100)))
pred <- predict(m2_ck, newdata = pdat, type = "link", se.fit = TRUE)
ilink <- family(m2_ck)$linkinv
pdat <- pdat %>%
  bind_cols(data.frame(pred)) %>%
  mutate(fitted = ilink(fit),
         upper = ilink(fit + (1.96 * se.fit)),
         lower = ilink(fit - (1.96 * se.fit)))

ggplot(ck, aes(x = ck, y = proportion)) +
  geom_ribbon(data = pdat,
              aes(ymin = lower, ymax = upper, x = ck),
              alpha = 0.2, inherit.aes = FALSE) +
  geom_point() +
  geom_line(data = pdat, aes(y = fitted, x = ck)) +
  labs(x = "Creatinine kinase level", y = "Proportion Heart Attack")
```
   
    